---
This tutorial was created by Eric Gilliam (egillia3@alumni.stanford.edu) for Dr. Steven Levitt's class "Data Construction and Interpretation 
in Economic Applications", a class he TA'd, offered at the University of Chicago in the Spring of 2021. It is included, without edit, at his courtesy.

title: "Econ 213 RStudio Tutorial: Where it pays to go to school"
author: "Eric Gilliam"
email: "egillia3@alumni.stanford.edu"
output: html_document
---
## Downloading and installing R and Rstudio

First things first, you need to download R for your computer. Download the 
appropriate mac or windows version here:
https://cran.rstudio.com/

Next, you need to download Rstudio. Download the appropriate mac or windows 
version here:
https://rstudio.com/products/rstudio/download/#download

I'm pretty sure you don't have to install xquartz, but if in the 
end you can't knit install it from here:
https://www.xquartz.org/. But just don't worry about this until they very end
of the tutorial


# Learning Some of the Basics

If you're reading this, then it means that you've successfully installed 
RStudio, which, frankly, is pretty much half the battle when it comes to 
learning R.

We're going to do a little tutorial to learn some of the basics in R. Now,
like any coding language, no matter how much is covered in this tutorial, you'll
run into a new problem/bug like every third line. And, like in every other
coding language, the best way to solve the problem is usually some combination
of Googling, asking a friend, and repeatedly trying again until the gods of R
have decided that you've had enough suffering. 

The document we're in right now is called an Rmarkdown document. In these
documents, you can write text outside of the code chunks (where you're reading
now) and write code in what are called code chunks. You'll see what a code
chunk is in a second.

## Installing and Loading Libraries

The first thing we will do is install and load in all of the required packages.
You only need to install a package once on each computer, but you will need to
load the packages at the start of each R session. (you install using 
install.packages() and load using library())


First things first, run each of the following 4 lines of code one by one in the
console one by one:
  - install.packages("ggplot2")
  - install.packages("tidyverse")
  - install.packages("stringr")
  - install.packages("devtools)

If you don't know how to do this, here's the step by step guide:
  - make sure the console tab is open at the bottom of your RStudio screen
  (The options are Console, Terminal R markdown, and Jobs). Click Console
  - copy the line of code from above (one at a time)
  - paste the line of code next to the '>' character
  - hit enter
  - things should happen in the console (don't worry about what they mean)
  - if it prompts you to answer a question, probably just type 'Yes' and hit enter
  - if something goes wrong, then google it and do whatever google tells you to
  do, but the odds are nothing will go wrong
  
Once this is done, do the same thing with install.packages("glue") and
install.packages("fs"), but for these two when it prompts you to write Yes
or no, write no.
  

The above lines are how you install packages. If you do already have them 
installed, then I think this will just update them or reinstall them. If you have
them installed already, then you don't have to run them probably

Below, the thing that starts with these things ``` is called a code chunk, 
this is where you will write your code. The keyboard short cut to create a new 
code chunk in a Mac is control + option + i. The keyboard shortcut for a PC is 
control + alt + i (I think).

From now on, we will not be running lines of code in the console, but, rather,
inside the code chunks. To run the lines of code in the code chunks, you have 
three different options:
  (1) You can click the green arrow in the top right of the chunk to run the entire 
      chunk at once
  (2) highlight a group of lines and hit command + enter if you just
      want to run that subset of lines
  (3) put your cursor on a line and hit command enter if you just want to run an 
      individual line

Below, we will load the packages we installed. You need to load packages at the
start of each R session.

```{r}
## Everybody should run these lines now matter what
library(ggplot2)
library(tidyverse)
library(stringr)




## The lines that start with '##' are known as comments. Comments are how you
## write text and notes to yourself from within a code chunk. If a line starts
## with '##', then RStudio knows to ignore that line/text when compiling code
```

## A Note

I used my sister, Serena, as a guinea pig for this tutorial. And from that 
experience I'll leave you with four pieces of advice/factoids that will help you
get through this tutorial:
  (1) Run every line of code in all the code chunks (I thought that was obvious)
  (2) Even if you don't know what something means, still run the line of code 
      (as per (1))
  (3) If you don't know what something means or get confused, you can probably
      just run the line and hope that it'll sink in later on
  (4) a 'Kiwi' is a person from New Zealand

## Reading in Datasets

The next thing we are going to do is read in a csv into R. To do this, first
go to this link and download the salaries-by-college-type.csv dataset:
https://www.kaggle.com/wsj/college-salaries/version/1#salaries-by-college-type.csv

We are going to write the path to the file in your computer in quotes in the 
command below. The path below is the path to the file in my computer as an 
example.

We will be using the function read_csv because the object we are trying to
read in is a .csv. There is also a read_dta function in the haven package for
reading in state files and a read_excel function in the read_xl package if
you want to read in xlsx files. If you didn't understand any of that, just 
don't worry about it for now

```{r}
##run this line, but change the file path to be the path in your computer, not 
## mine
read_csv("/Users/ericmatthewgilliam/Downloads/salaries-by-college-type.csv")


```

If the above line of code worked, then there should have been a green flash
on the left side of the chunk and some snapshots of the data should have 
appeared below the code chunk for you to look through. It's like magic

Now, here's the thing, if you want any object in R to continue to exist in
a way that you can reference it/use it later, then you will have to 'assign' it
a name. 

To do this, you just put a name you've picked at the start of the line,
followed by an arrow '<-' and then write the line of code which creates
the object, variable, or whatever it is you're creating

```{r}
##run this line to load in the dataset and save it using the name college_data
## Again, change the file path to the path in your computer
college_data <- read_csv("/Users/ericmatthewgilliam/Downloads/salaries-by-college-type.csv")

##Now that you've run this line, in the top right of your screen, under Global
##Environment, click on college_data and you should be able to view the whole 
## dataset in R

##another way to do this is to run the line View(college_data) in the console


```

## Getting to know your data

First , let's get to know our data a bit.

```{r}
## Run the following line to see how many rows are in the data (how many colleges
## are in the dataset)
nrow(college_data)
```


```{r}
##Run the following line to see your column names
colnames(college_data)
```


```{r}
##Run the following line to see all of the different unique values in the 
## School type column
unique(college_data$`School Type`)

## For reference, if you simply write the name of the dataset followed by a
## dollar sign such as college_data$ , a drop down menu should appear with
## all of the columns in the dataset. 

unique(college_data$`Starting Median Salary`)

```

Before we start graphing our data, let's take a look at each of the columns
to make sure that the columns are of the type we believe they are

```{r}
## Run the following line of code to ensure that your columns of numbers 
## are being stored as integers or double (numeric types) rather than as
## characters. 
typeof(college_data$`Starting Median Salary`)
```


```{r}
## It appears that the median salary column is being stored as a character

## run the following line to see if all of the columns that should be numbers
## are being stored a characters
glimpse(college_data)
```

Let's change these columns from character types to doubles. A type 'double'
means a column is a numeric type with decimals

I'm not going to get in to what are strings vs booleans vs doubles vs integers
in this tutorial, but just know that the reason we have to change the strings
that look like numbers to actual numbers is because, to the computer, trying 
to add together any two strings, even if they look like numbers, is like telling
it to add together the words "Eric" and "cool". Like yeah, it makes sense to 
you because you get it, but the computer does not. I checked.

I will show two ways to make new columns

But first I'm gonna remove the dollar signs from the columns. You can 
ignore this next chunk of code

Run this next chunk of code by clicking the green arrow, but don't worry about
what's happening

```{r}
##ignore
college_data <-
  college_data %>% 
  mutate(
    `Starting Median Salary` = 
      str_replace(`Starting Median Salary`, c('\\$'), ""),
    `Mid-Career Median Salary` = 
      str_replace(`Mid-Career Median Salary`, '\\$', ""),
    `Mid-Career 10th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\$', ""),
    `Mid-Career 25th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\$', ""),
    `Mid-Career 50th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\$', ""),
    `Mid-Career 75th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\$', ""),
    `Mid-Career 90th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\$', ""),
    `Starting Median Salary` = 
      str_replace(`Starting Median Salary`, c('\\,'), ""),
    `Mid-Career Median Salary` = 
      str_replace(`Mid-Career Median Salary`, '\\,', ""),
    `Mid-Career 10th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\,', ""),
    `Mid-Career 25th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\,', ""),
    `Mid-Career 50th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\,', ""),
    `Mid-Career 75th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\,', ""),
    `Mid-Career 90th Percentile Salary` = 
      str_replace(`Mid-Career 10th Percentile Salary`, '\\,', ""),
  )
```

Okay now I'll actually show you how to make new columns

Below we will change the columns that look like numbers, but are actually
characters, to actually be numbers


```{r}
## The first way to make a new column is using base R. To make new columns in 
## base R you write the name of the dataset, followed by the dollar sign,
##  followed by what you want the new column to be called and an arrow
## then you perform the operation to create the column

## We will make a new column named starting_median_salary below
college_data$starting_median_salary <- as.double(college_data$`Starting Median Salary`)

##running this line of code prints the column below
college_data$starting_median_salary
##looks good
```



What we just did was make a completely new column. In reality, the 
old columns that are listing numbers as strings are useless to us. So,
instead of making brand new columns, we will replace the old ones with the new
ones. The best way to do this is to make a new column that has the exact same
name as the old column. This effectively overwrites the old column with the 
new column

I'll demonstrate this to you using the second method of column creation 
which is creating columns using the tidyverse. This means we will be using
pipe operators ( %>% ) which can be typed using control + shift + m
and the mutate function

Using the tidyverse and pipe operators frees you from constantly having to
type the name of the dataset followed by the dollar sign. Instead, you
start your lines with the name of the dataset followed by the pipe operator,
after that you do not have to write the name of the dataset again, just the
column names. Also, it's  much easier to read, which makes it easier for you
to de-bug your code when you inevitably run in to problems

The mutate function creates new columns, and I will use that below
Below, we create the new columns of type double and save the dataset
After this, this dataset should be ready to play with

If some of this doesn't make sense yet, just roll with it and some of it will
sink in as you go on. You'll start to see the patterns

```{r}
college_data <-
  college_data %>% 
  mutate(
    `Starting Median Salary` = as.double(`Starting Median Salary`),
    `Mid-Career Median Salary` = as.double(`Mid-Career Median Salary`),
    `Mid-Career 10th Percentile Salary` = as.double(`Mid-Career 10th Percentile Salary`),
    `Mid-Career 25th Percentile Salary` = as.double(`Mid-Career 25th Percentile Salary`),
    `Mid-Career 50th Percentile Salary` = as.double(`Mid-Career 50th Percentile Salary`),
    `Mid-Career 75th Percentile Salary` = as.double(`Mid-Career 75th Percentile Salary`),
    `Mid-Career 90th Percentile Salary` = as.double(`Mid-Career 90th Percentile Salary`)
  )

glimpse(college_data)

##Since we made two starting median salary columns by different names, you can
## remove the first one by doing this

college_data <-
  college_data %>% 
  select(-starting_median_salary)

##If you did't understand how the above line worked, just move on and don't 
## worry about it for now

## and if you're a perfectionist who refuses to move on until you understand
## exactly how every single piece of code you run works backwards and forwards,
## then coding is going to be a very frustrating experience for you
```

If at any point you really need/want to know how a function works. Just write
the name of the function preceded by a '?' and run the line, and the RDocumentation
for that function should show up in the bottom right quadrant of your RStudio
screen

For example, run the following line of code below.
```{r}
?read_csv
```

Some descriptions of how the function works should have appeared on the right
side of your screen. If after looking at this documentation you are still 
confused, Google is your best friend. 

## First things first, I want to know what the party schools are

Before we move on, we all probably want to know what the Wall Street
Journal classified as the 'Party' schools. So I'll show you how to do that

We use the filter function, which filters out all of the rows that don't match
the given condition

```{r}
party_schools <-
  college_data %>% 
  filter(`School Type` == "Party") ## in R you use == to check if two things are equal

party_schools

```

Now, I know what you're all thinking, "How in the hell is UChicago not on there?!"

Well, I looked into it, and UChicago isn't in the dataset as a whole. I think
it's safe to say you'd have made it into the 'Party' school category if you
were included. 

Like all the kids at UT Austin would tell you, "Nothing makes a good party like
some rigorous Marx discussion with indie folk music playing at a tasteful
volume in the background."

## Basic Graphics and Data Manipulation in R

We are going to use tidyverse and ggplot to make some basic visualizations
to explore your data. Ggplot and tidyverse are the best and you should 
use them. The guy who wrote them, Hadley Wickham, just won the equivalent
of the Nobel Prize in statistics, is a Kiwi, and has phenomenal taste in 
earrings. So, you should obviously do what he says.

Below, I am going to show you a bunch of different examples of basic graphics
and data manipulations so you can get the hang of how things are done and which
you can refer back to for some basic guidance.

You might not understand exactly how all of the lines being run below work,
and that's fine. I could explain it all, but it would make the tutorial 
convoluted and not very focused/efficient. Just make sure you get the
gist of how these work. The goal is to show you what's possible and roughly 
how to mimic the code below to make your own graphs even if you don't 
understand why everything works. 

If you want to know more, you can just come to my office hours.  

### Histogram
Let's take a look at the distribution of starting median salaries
in the dataset

We will make these graphs using ggplot and geom_histogram

```{r}
college_data %>% 
  ggplot(aes(x = `Starting Median Salary`)) +
  geom_histogram()
```

Note, after ggplot lines you use plus signs as opposed to pipe operators (%>%),
there's no particular reason for this. I think one time Hadley told my class
that the only reason this is the case is because he wrote ggplot first and he
didn't feel like going back and changing it after he'd worked on the tidyverse.

## Scatter Plot

Let's make a scatter plot of starting median salaries and median mid career
salaries

We will make this using geom_point

```{r}
college_data %>% 
  ggplot(aes(x = `Starting Median Salary`, y = `Mid-Career Median Salary`)) +
  geom_point()
```

Unsurprisingly these are extremely highly correlated. Let's fit a quick curve
to this data using geom_smooth

```{r}
college_data %>% 
  ggplot(aes(x = `Starting Median Salary`, `Mid-Career Median Salary`)) +
  geom_point() +
  geom_smooth()
```


### Ordering Tables


Below, let's arrange a table from highest starting salaries to lowest
starting salaries

We will do this using the arrange function
```{r}
##This orders the rows from lowest to highest
college_data %>% 
  arrange(`Starting Median Salary`)

##to reverse this just put a minus sign in front starting median salary
college_data %>% 
  arrange(-`Starting Median Salary`)
```


```{r}
## if you would like to just look at a subset of columns, then use the select
## function and pick the columns you'd like to see
college_data %>% 
  arrange(-`Starting Median Salary`) %>% 
  select(`School Name`, `Starting Median Salary`)
```


```{r}
## if you'd like to only see the top 20 schools, once you've used arrange,
## you can use the slice function
college_data %>% 
  arrange(-`Starting Median Salary`) %>% 
  select(`School Name`, `Starting Median Salary`) %>% 
  slice(1:20)

## To save all of the information for the top 20 schools just use the assignment
## operator

top_20_schools <-
  college_data %>% 
  arrange(-`Starting Median Salary`) %>% 
  slice(1:20)


```



## Grouping Tables

What if we wanted to Look at the mean/median incomes for each school type?

The best way to do this is using the group_by() and summarise() functions in the
tidyverse

The group_by operator indicates that you want a certain operation done only
on a specific grouping of rows, such as wanting the median FOR EACH type
of school

Look at the output of the code below
```{r}
college_data %>% 
  group_by(`School Type`) %>% 
  summarise(median_starting_salary = median(`Starting Median Salary`))
```

group_by() tells the data that you want to only have the operation done using
schools in the same group

summarise() creates a new column that compresses all of the data points in a 
given group, such as party schools, to a single value (such as the mean or sum
of a column for that group)


Below we will make a similar table, but including the median starting
salary, the standard deviation of those values, the median mid career
salary, and the standard deviation of those values

```{r}
college_data %>% 
  group_by(`School Type`) %>% 
  summarise(
    median_starting_salary = median(`Starting Median Salary`),
    sd_starting_salary = sd(`Starting Median Salary`),
    median_midcareer_salary = median(`Mid-Career Median Salary`),
    sd_midcareer_salary = sd(`Mid-Career Median Salary`)
    )
```

Below I will make the same table, but using means instead of medians

```{r}
college_data %>% 
  group_by(`School Type`) %>% 
  summarise(
    mean_starting_salary = mean(`Starting Median Salary`),
    sd_starting_salary = sd(`Starting Median Salary`),
    mean_midcareer_salary = mean(`Mid-Career Median Salary`),
    sd_midcareer_salary = sd(`Mid-Career Median Salary`)
    )
```

Now, let's say you wanted to look at mid career 10th percentile salaries
to assess the worst case scenarios of attending a particular type of college

We run into some complications. We get mostly NA's. Why is this?

```{r}
college_data %>% 
  group_by(`School Type`) %>% 
  summarise(
    tenthpct_starting_salary = mean(`Mid-Career 10th Percentile Salary`),
    sd_tenthpct_salary = sd(`Mid-Career 10th Percentile Salary`)
    )
```


If you go look at the Mid-Career 10th Percentile Salary column in the dataset,
you will see why. There are NA's in the column. NA's occur where no data was
collected. If you try to perform an  operation on a column or vector with NA's, 
this can tend to happen. R does not know how to include the NA in the average. 

There are no NA's in mathematics, so that's something you have to use your own 
judgement to take care of. Now, I know a lot of you might not trust  
your judgement very much/if at all. But even if you're right in doubting your
judgement, in these cases you kind of just have to make a call. In this scenario
there are some common options:
  
  - make all NA's 0's (probably not a good call in this case)
  
  - calculate the mean of the column after removing those rows with NAs
  
  - replacing all of the NAs with the mean of the non NAs in the column
  

On the bright side, even if you get it wrong, in the real world people usually 
don't check and see what you did. Now, it's probably best for your grade in this
class and the University of Chicago's image that you not make too many 
boneheaded mistakes. But, if we're being honest with each other, most people
don't really know or care how a linear regression works. And if they don't care
about that, then they probably REALLY don't care about what you did with your
NA's. BUT, if someone ever does spot your mistake, then everything else you've
shown in your analysis gets called into question. And that means you've wasted
your time coding in RStudio instead of going out and being a cool kid for
nothing. 

Here's a funny House of Lies clip on the subject if you have nothing better to
do: 
https://www.youtube.com/watch?v=4cyvmF7GD8o (start at 3:00)

Long story short: don't waste your own time, do the little things right

Now, below I'll show you how to take the mean by removing the NA's. If you'd 
like to know how to do this using median, just replace the mean function with median
and you're in business, and if you'd like to know how to do this by replacing 
NAs with the mean, you can  google it. It's called 'mean imputation'. Have fun

```{r}
college_data %>% 
  group_by(`School Type`) %>% 
  summarise(
    tenthpct_starting_salary = mean(`Mid-Career 10th Percentile Salary`, na.rm = TRUE),
    sd_tenthpct_salary = sd(`Mid-Career 10th Percentile Salary`, na.rm = TRUE)
    )
```

Now, you're probably wondering, "Eric, can I graph this new data too?!"

Yeah...you can graph everything and you can name everything (for the most part)

```{r}
college_data %>% 
  group_by(`School Type`) %>% 
  summarise(
    tenthpct_starting_salary = mean(`Mid-Career 10th Percentile Salary`, na.rm = TRUE),
    sd_tenthpct_salary = sd(`Mid-Career 10th Percentile Salary`, na.rm = TRUE)
    ) %>% 
  ggplot(aes(x = `School Type`, y = tenthpct_starting_salary)) +
  geom_point()
```

Now, we're getting towards the end here. I hope you've been having as much fun
as I have. And if you haven't then you're more than welcome to go do some
random online R tutorial. I get paid regardless. Actually... I'm not even sure
if I get paid. I haven't asked Steve

Anyways, let's put together some of the skills we learned above and make a new
column which calculates the increase between median starting salary and 
median mid career salary for each school. We'll call this column
poor_today_rich_tomorrow

```{r}
college_data <- 
  college_data %>% 
  mutate(poor_today_rich_tomorrow = `Mid-Career Median Salary` - `Starting Median Salary`)
```

Let's look at the distribution of these scores. It's good practice to always
do this so you get a sense of your variables. A lot of economists are bad about doing
this, but try to be better than them. Make them insecure (more insecure)

```{r}
college_data %>% 
  ggplot(aes(poor_today_rich_tomorrow)) +
  geom_histogram()
```


Now, let's make a graph so big it won't even fit on the screen. That should
be fun and not at all useless.

We're gonna make a graph where each point represents a single schools 
poor_today_rich_tomorrow score

```{r, fig.height=40, fig.width=10}
##fct_reorder reorders the school names column by the rich_poor score
##this makes the graph look prettier
##if you don't understand it don't worry about it
college_data %>% 
  mutate(`School Name` = fct_reorder(`School Name`, poor_today_rich_tomorrow)) %>% 
  ggplot(aes(`School Name`, poor_today_rich_tomorrow)) +
  geom_point() +
  coord_flip()
  
```

This graph will look ugly here, but will look pretty and be helpful when it
shows up in the knit html file at the end of the tutorial.

Let's calculate the average poor_today_rich_tomorrow score for each type of school
below

```{r}
college_data %>% 
  group_by(`School Type`) %>% 
  summarise(
    mean_income_growth = mean(poor_today_rich_tomorrow, na.rm = TRUE),
    sd_income_growth = sd(poor_today_rich_tomorrow, na.rm = TRUE)
    )
```

Below, I'll make a graph with a title and labels from the data above

```{r}
college_data %>% 
  group_by(`School Type`) %>% 
  summarise(
    mean_income_growth = mean(poor_today_rich_tomorrow, na.rm = TRUE),
    sd_income_growth = sd(poor_today_rich_tomorrow, na.rm = TRUE)
    ) %>% 
  ggplot(aes(x = `School Type`, y = mean_income_growth)) +
  geom_histogram(stat = "identity") +
  labs(
    title = "Must be Nice to go to an Ivy",
    y = "Income Growth",
    caption = "'But Mom, U of C is kind of like the Ivy League'"
  )
```


## Regressions

The last thing I'll show you how to do is fit a basic regression. 

Below, we'll fit a multiple linear regression which attempts to use a school's
median starting salary and the school type to predict a school's median
mid-career salary

The notation for this is pretty simple. 

```{r}
##This line runs the regression and saves it in linear_model
linear_model <- lm(`Mid-Career Median Salary` ~ `Starting Median Salary` + `School Type`, data = college_data)

##this line prints out the summary of your regression
summary(linear_model)

```

The low p values in the 'Pr(>|t|)' column and those asterisks which the 
economists love so much means that all of those variables above are 
statistically significant and useful in predicting your mid career salary. The 
'Estimate' column shows the coefficient of each variable in the multiple linear
regression. The lm function which we just used can be used to fit all sorts 
of regressions and curves and will probably be useful in your next assignment

If you have any questions about this tutorial, the class, R coding in general,
my favorite color (baby blue), or my hopes and dreams feel free to email me 
at ericgilliam@uchicago.edu and I should get back to you pretty quickly OR you 
can just drop into my office hours. 

## In Conclusion

Using RStudio is great because you can 'knit' the output of your file to html,
pdf, or some other kind of file. Then your final analysis can be in a pretty,
more easy to read form than this Rmarkdown file. The last step of this tutorial
is to go to the top of the Rmarkdown screen and hit the 'Knit' button. It should
knit this file to an html document that you can scroll through. If you want, you
can save the html file and send it to people. Like, my mom would never care if I
sent it to her (she doesn't even know what I do for a living), but maybe someone 
in your life will care. idk


.


.


.

.

.

.

.

.

